A standardized framework for risk-based assessment of treatment effect heterogeneity in observational healthcare databases

### 2022-06-02 submission/manuscript.rmd
prepare manuscript submission

---
title: "A standardized framework for risk-based assessment of treatment effect heterogeneity in observational healthcare databases"
author: 
  - Alexandros Rekkas, MSc${^1}$ 
  - David van Klaveren, PhD$^{2,3}$, 
  - Patrick B. Ryan, PhD$^4$
  - Ewout W. Steyerberg, PhD$^{3,5}$
  - David M. Kent, PhD$^2$
  - Peter R. Rijnbeek, PhD${^1}$
output:
    bookdown::pdf_document2: default
    bookdown::word_document2:
        reference_docx: reference.docx
geometry: margin=1.0in
toc: false
font-size: 11pt
header-includes:
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage[left, pagewise]{lineno}
  - \usepackage{caption}
editor_options: 
  chunk_output_type: console
bibliography: references.bib
csl: jamia.csl
---

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(knitr)
library(here)
library(kableExtra)
d <- function(x, decimals = 2) {
  sprintf(paste0("%1.", decimals, "f"), x) 
}
inline_hook <- function(x) {
  if (is.numeric(x)) {
    if (abs(x - round(x)) < .Machine$double.eps) {
      formatted <- format(x, digits = 1, big.mark = ",")
    } else {
      formatted <- format(x, digits = 1 ,nsmall = 2, big.mark = ",")
    }
  } else {
    formatted <- x
  }

@@

knit_hooks$set(inline = inline_hook)
```
\thispagestyle{empty}
\vspace{8mm}

@@

This work has been performed in the European Health Data and
Evidence Network (EHDEN) project. This project has received funding from the
Innovative Medicines Initiative 2 Joint Undertaking (JU) under grant agreement
No 806968. The JU receives support from the European Union’s Horizon 2020
research and innovation programme and EFPIA.

\newpage
\newpage
# Abstract {-}
\singlespacing 
One of the aims of the Observational Health Data Sciences and Informatics
(OHDSI) initiative is population-level treatment effect estimation in large
observational databases. Since treatment effects are well-known to vary across
groups of patients with different baseline risk, we aimed to extend the OHDSI
library of open-source tools with a framework for risk-based assessment of
treatment effect heterogeneity. The proposed framework consists of five
steps: 1) definition of the problem, i.e. the population, the treatment, the
comparator and the outcome(s) of interest; 2) identification of relevant
databases; 3) development of a prediction model for the outcome(s) of
interest; 4) estimation of relative and absolute treatment effect within strata
of predicted risk, after adjusting for observed confounding; 5) presentation of
the results. We demonstrate our framework by evaluating heterogeneity of the
effect of angiotensin-converting enzyme (ACE) inhibitors versus beta blockers on
3 efficacy and 6 safety outcomes across three observational databases. Patients
    at low risk of acute myocardial infarction (MI) received negligible absolute
benefits for all 3 efficacy outcomes, though they were more pronounced in the
highest risk quarter, especially for hospitalization with heart failure. The
substantial risk increase of cough and angioedema with ACE inhibitors across all
risk strata suggests that beta blockers provide a viable alternative to patients
at low risk of acute MI. The proof of concept study demonstrates its feasibility
in large observational data. Further insights may arise by application to safety
and effectiveness questions across the global data network.

@@

The Observational Health Data Science and Informatics (OHDSI) collaborative has
established a global network of data partners and researchers that aim to bring
out the value of health data through large-scale analytics by mapping local
databases to the Observational Medical Outcomes Partnership (OMOP) Common Data
Model (CDM) [@hripcsak2015observational;@Overhage2012]. A standardized framework
applying current best practices for comparative effectiveness studies within the
OHDSI setting has been proposed [@Ryan2013]. This framework was implemented on a
large scale for estimation of average effects of all first-line hypertension
treatment classes on a total of 52 outcomes of interest across a global network
of 9 observational databases [@Suchard2019].

Treatment effects can often vary substantially across individual patients,
causing overall effect estimates to be inaccurate for a significant proportion
of the patients at hand [@Rothwell1995;@KRAVITZ2004]. Understanding
heterogeneity of treatment effects (HTE) has been central to the agenda for both
personalized (or precision) medicine and comparative effectiveness research,
giving rise to a wide range of approaches for its discovery, evaluation and
application in clinical practice. While exploratory and confirmatory HTE
analyses focus on generating and testing hypotheses on subgroup effects,
predictive HTE analyses use all available patient data to predict the benefits
or harms of treatment in individual patients [@Varadhan2013;@Kent2018].

Baseline risk has been recognized as a robust predictor of treatment effect,
because of the inherent relationship between baseline risk and the absolute
effects of treatment
[@Rothwell2005;@Hayward2006;@Kent2007;@Kent2008;@Kent2010]. For example, an
invasive coronary procedure – in comparison with medical treatment – improves
survival in patients with myocardial infarction at high (predicted) baseline
risk but not in those at low baseline risk [@Thune2005]. It has also been shown
that high-risk patients with pre-diabetes benefit substantially more from a
lifestyle modification program than low-risk patients [@Sussman2015].

Recently, systematic guidance on the application of risk-based methods has been
developed for RCT data [@Kent2019;@PathEnE]. Patients are divided into risk strata using
either an existing or an internally developed risk prediction
model. Risk-stratum-specific estimates provide an overview of the evolution of
treatment effects with increasing risk both on the relative and the absolute
scale. Several methods for predictive HTE analysis have been adapted for use in
observational data, but risk-based methods for predictive HTE in observational
data are still lacking and have been highlighted as an important future research
need [@PathEnE].

We aimed to develop a framework for risk-based assessment of treatment effect
heterogeneity in observational healthcare databases, extending the existing
methodology developed for the RCT setting. We implemented the framework in a
publicly available package providing an out-of-the-box solution for implementing
such analyses at scale within any observational database mapped to OMOP-CDM. In
a case study we analyzed heterogeneity of the effects of first-line hypertension
treatment: we compared the effect of angiotensin converting enzyme (ACE)
inhibitors to beta blockers on 9 outcomes across three different US claims
databases.

@@

The proposed framework defines 5 distinct steps that enable a standardized
approach for risk-based assessment of treatment effect heterogeneity for
databases mapped to the OMOP-CDM. These are: 1) general definition of the
research aim; 2) identification of the databases within which the analyses will
be performed; 3) a prediction step where internal or external prediction models
are used to assign patient-level risk predictions; 4) an estimation step where
absolute and relative treatment effects are estimated within risk strata; 5)
presentation of the results. We developed an open-source R-package for the
implementation of the proposed framework and made it publicly available. The
source code can be found at https://github.com/OHDSI/RiskStratifiedEstimation.

@@

Our framework uses a comparative cohort design. This means that at least 3
cohorts of patients need to be defined at this stage of the framework:

@@

Including in our analyses multiple databases representing the population of
interest potentially increases the generalizability of results. Furthermore, the
cohorts should preferably have adequate sample size with adequate follow-up time
to ensure precise effect estimation, even within smaller risk strata. Other
issues that may be of importance for database inclusion are the depth of data
capture (the precision at which measurements, lab tests, conditions are
recorded) and the reliability of data entry.

@@

We adopt the standardized framework for the generation of patient-level
prediction models using observational data that ensures adherence to existing
guidelines [@Collins2015;@Moons2015]. To generate the target cohort (i.e. the
set of patients on which the prediction model will be developed), we pool the
already defined treatment cohort and comparator cohort. To avoid deferentially
fitting the prediction model to patients across treatment arms, thus introducing
spurious interactions with treatment [@Burke2014;@vanKlaveren2019], we develop
the patient-level prediction model in the propensity score-matched (1:1) subset
of the sample. Finally, we need to define the time horizon within which we aim
to make predictions and we also need to select the machine-learning algorithm we
want to use to generate patient-level predictions.

After model development, a performance overview of the derived prediction models
including discrimination and calibration both in the propensity score matched
subset, the entire sample and separately for treated and comparator patients
should also be reported.

@@

We estimate treatment effects (both on the relative and the absolute scale)
within risk strata—typically 4 risk quarters—defined using the prediction model
of step 3. Effect estimation may be focused on the difference in outcomes for a
randomly selected person from the risk stratum (average treatment effect) or for
a randomly selected person from the treatment cohort within the risk stratum
receiving the treatment under study (average treatment effect on the treated).

@@

Prior to analyzing the results, we need to assess if adequate covariate balance
was achieved within each risk stratum accounting for measured
confounding. Common approaches include analysis of the overlap of propensity
score distributions and calculation of standardized mean differences of the
covariates before and after propensity score adjustment.

A schematic overview of the prediction and estimation steps is presented in
Figure XX.

@@

In the presence of a positive treatment effect and a well-discriminating
prediction model we expect an increasing pattern of the differences in the
absolute scale, even if treatment effects remain constant on the relative scale
across risk strata. Due to this scale-dependence of treatment effect
heterogeneity, results should be assessed both on the relative and the absolute
scale. We find that a side-by-side presentation on a forest-like format can give
a very good representation of our results.

@@

As a demonstration of our framework, we evaluated if our proposed method was
able to identify treatment effect heterogeneity of ACE inhibitors compared to
beta blockers using acute myocardial infarction (MI) risk quarter specific
effect estimates, both on the relative and on the absolute scale. We focused on
3 efficacy outcomes (acute MI, hospitalization with heart failure and ischemic
or hemorrhagic stroke) and 6 safety outcomes (hypokalemia, hyperkalemia,
hypotension, angioedema, cough and abnormal weight gain). We used data from 3
US-based claims databases. The analysis plan was the framework outlined in
steps 1 through 5.

A recent meta-analysis has shown that beta-blockers are on average less
efficacious in preventing cardiovascular events than ACE inhibitors
[@Wiysonge2017]. However, meta-analyses focusing on age of the patients have
shown that beta blockers have similar efficacy in younger hypertensive patients,
but should not be used as first-line treatment of hypertension in older patients
mainly due to their increased risk of cardiovascular events compared to other
first-line antihypertensive treatments [@Khan2006;@Vgele2017].Our aim was to
replicate these conclusions in our analyses.

@@

We considered the following research aim: “compare the effect of ACE-inhibitors
($T$) to the effect of beta blockers ($C$) in patients with established hypertension
with respect to 9 outcomes ($O_1,\dots,O_9$)”. The cohorts are:

- Treatment cohort: Patients receiving any drug within the ACE-inhibitor class
  with at least one year of follow-up before treatment initiation and a recorded
  hypertension diagnosis within that year.
- Comparator cohort: Patients receiving any drug within the beta blocker class
  with at least one year of follow-up before treatment initiation and a recorded
  hypertension diagnosis within that year.
- Outcome cohorts: We considered 3 efficacy and 6 safety outcome cohorts. These
  were patients in the database with a diagnosis of: acute MI; hospitalization
  with heart failure; ischemic or hemorrhagic stroke (efficacy outcomes);
  hypokalemia; hyperkalemia; hypotension; angioedema; cough; abnormal weight
  gain (safety outcomes).

@@

For our analyses we used data from 3 US claims databases, namely IBM MarketScan
Commercial Claims and Encounters (CCAE), IBM MarketScan Medicaid (MDCD), and IBM
MarketScan Medicare Supplemental Beneficiaries (MDCR). Our analyses included a
total of 784,561, 66,820 and 101,661 patients initiating treatment with ACE
inhibitors and 395,740, 45,999 and 69,798 patients initiating treatment with
beta blockers in CCAE, MDCD and MDCR respectively (Table 1). Adequate numbers of
patients were included in all strata of predicted acute MI risk.

@@

We internally developed separate prediction models for acute MI in all 3
databases. The prediction models were estimated on the propensity score matched
(1:1) subset of the sample, using caliper of 0.2 and after excluding patients
having the outcome any time prior to treatment initiation. We chose a time
horizon of 2 years after inclusion into the target cohort. We developed the
prediction models using LASSO logistic regression with 3-fold cross validation
for hyper-parameter selection.

The models developed in the 3 databases had moderate discriminative performance
(internally validated) with no major issues of overfitting to any cohort except
for the case of CCAE database in which the derived prediction model performed
better in the comparator cohort (Table 2). We also observed lower performance of
the prediction model developed in MDCR compared to the other 2 databases (see
supplementary material).

@@

Our aim was to estimate the average treatment effects on the relative and the
absolute scale within strata of predicted acute MI risk.

We used patient-level predictions to stratify the sample into 4 MI risk
quarters. Within risk quarters, relative effects were estimated using Cox
regression and absolute effects were estimated from the Kaplan-Meier estimate
differences at 2 years after treatment initiation. To adjust for observed
confounding within risk strata, we estimated propensity scores using the same
approach as in the development of prediction models. We used the estimated
propensity scores to stratify patients into 5 strata, within each risk
quarter. The risk quarter-specific effect estimates were derived by averaging
over the estimates within the propensity score fifths.

There was sufficient overlap of propensity score distribution in all risk strata
(Figure 2). Propensity score adjustment achieved balance for most of the
considered covariates, measured using standardized mean differences before and
after adjustment (Figure 3). However, in lower risk strata imbalances persisted
for a substantial subset of the covariates, all related to pregnancy
findings. This was anticipated (use of ACE inhibitors is specifically
contraindicated during pregnancy) and was already pointed out in
[@Suchard2019]. Therefore, treatment effect estimates in the lowest risk
quarters of CCAE and MDCR may still be subject to residual confounding and
should be interpreted with care.

@@

```{r, echo=FALSE}
map_outcomes <- readRDS(here("data/processed/multipleRseeAnalyses/map_outcomes.rds"))
map_exposures <- readRDS(here("data/processed/multipleRseeAnalyses/map_exposures.rds"))
relative <- readRDS(here("data/processed/multipleRseeAnalyses/mappedOverallRelativeResults.rds")) %>%
  left_join(map_outcomes, by = c("estOutcome" = "outcome_id")) %>%
  select(-estOutcome) %>%
  rename("estOutcome" = "outcome_name") %>%
  left_join(map_outcomes, by = c("stratOutcome" = "outcome_id")) %>%
  select(-stratOutcome) %>%
  rename("stratOutcome" = "outcome_name") %>%
  left_join(map_exposures, by = c("treatment" = "exposure_id")) %>%
  select(-treatment) %>%
  rename("treatment" = "exposure_name") %>%
  left_join(map_exposures, by = c("comparator" = "exposure_id")) %>%
  select(-comparator) %>%
  rename("comparator" = "exposure_name")

absolute <- readRDS(here("data/processed/multipleRseeAnalyses/mappedOverallAbsoluteResults.rds")) %>%
  left_join(map_outcomes, by = c("estOutcome" = "outcome_id")) %>%
  select(-estOutcome) %>%
  rename("estOutcome" = "outcome_name") %>%
  left_join(map_outcomes, by = c("stratOutcome" = "outcome_id")) %>%
  select(-stratOutcome) %>%
  rename("stratOutcome" = "outcome_name") %>%
  left_join(map_exposures, by = c("treatment" = "exposure_id")) %>%
  select(-treatment) %>%
  rename("treatment" = "exposure_name") %>%
  left_join(map_exposures, by = c("comparator" = "exposure_id")) %>%
  select(-comparator) %>%
  rename("comparator" = "exposure_name") %>%
  mutate(
    estimate = 100 * estimate,
    lower    = 100 * lower,
    upper    = 100 * upper
  )

@@

acute_cough_absolute <- absolute %>%
  filter(
    stratOutcome == "acute_myocardial_infarction",
    estOutcome == "cough"
  )
```

For hospitalization with acute MI there was an increasing trend in favor ACE
inhibitors compared to beta blockers on the relative scale (hazard ratios
decreased) with increasing acute MI risk. More specifically, hazard ratios
decreased from 

@@

95\% CI) in CCAE and MDCD respectively (Figure 4). In MDCR hazard ratios remained stable
from

@@

95\% CI). In terms of hospitalization with heart failure relative treatment
effect estimates favored ACE inhibitors across all risk strata in all
databases. We found no differences between the two treatments in their effect on
stroke on the relative scale. In terms of the safety outcomes, we found an
increased ACE-inhibitor risk of cough and angioedema on the relative scale
across all risk strata. In the case of cough, this effect decreased with
increasing risk of acute MI---from 

@@

## Evaluation

>>>>

---
title: "A standardized framework for risk-based assessment of treatment effect heterogeneity in observational healthcare databases"
author: 
  - Alexandros Rekkas, MSc${^1}$ 
  - David van Klaveren, PhD$^{2,3}$, 
  - Patrick B. Ryan, PhD$^4$
  - Ewout W. Steyerberg, PhD$^{3,5}$
  - David M. Kent, PhD$^2$
  - Peter R. Rijnbeek, PhD${^1}$
output:
    bookdown::pdf_document2: default
    bookdown::word_document2:
        reference_docx: reference.docx
geometry: margin=1.0in
toc: false
font-size: 11pt
header-includes:
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage[left, pagewise]{lineno}
  - \usepackage{caption}
editor_options: 
  chunk_output_type: console
bibliography: references.bib
csl: the-lancet.csl
---

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(knitr)
library(here)
library(kableExtra)
d <- function(x, decimals = 2) {
  sprintf(paste0("%1.", decimals, "f"), x) 
}
inline_hook <- function(x) {
  if (is.numeric(x)) {
    if (abs(x - round(x)) < .Machine$double.eps) {
      formatted <- format(x, digits = 1, big.mark = " ")
    } else {
      formatted <- format(x, digits = 1, nsmall = 2, big.mark = " ", decimal.mark = intToUtf8("0x00B7"))
    }
  } else {
    formatted <- x
  }

@@

knit_hooks$set(inline = inline_hook)
```
\thispagestyle{empty}

@@

This work has been performed in the European Health Data and
Evidence Network (EHDEN) project. This project has received funding from the
Innovative Medicines Initiative 2 Joint Undertaking (JU) under grant agreement
No 8069six. The JU receives support from the European Union’s Horizon 2020
research and innovation programme and EFPIA.

\newpage
\newpage
# Abstract {-}
\singlespacing 
**Background**: Treatment effects are often anticipated to vary across groups of
patients with different baseline risk. Implementation of a risk-based approach
in the RCT setting has shown evidence of increase in statistical power for the
evaluation of treatment effect heterogeneity. The aim of this study was to
extend this approach to the observational setting using a standardized scalable
framework.

**Methods**: The proposed framework consists of
five steps: 1) definition of the problem, i.e. the population, the treatment,
the comparator and the outcome(s) of interest; 2) identification of relevant
databases; 3) development of a prediction model for the outcome(s) of interest;
4) estimation of relative and absolute treatment effect within strata of
predicted risk, after adjusting for observed confounding; 5) presentation of the
results. We demonstrate our framework by evaluating heterogeneity of the effect
of angiotensin-converting enzyme (ACE) inhibitors versus beta blockers on three
efficacy and six safety outcomes across three observational databases.

**Findings**: In our demostration, patients at low risk of acute myocardial
infarction (MI) received negligible absolute benefits for all three efficacy
outcomes, though they were more pronounced in the highest risk quarter,
especially for hospitalization with heart failure. However, failing diagnostics
showed evidence of residual imbalances even after propensity score adjustment.

**Interpretation**: Application of our framework allows for
observing differential risk of outcomes across strata, which offers the
opportunity to consider the benefit-harm tradeoff between alternative treatments
within targeted subpopulations. Further insights may arise by application to
safety and effectiveness questions on a large scale.

**Funding**: European Health Data and Evidence Network (EHDEN) project.

@@

Treatment effects can often vary substantially across individual patients,
causing overall effect estimates to be inaccurate for a significant proportion
of the patients at hand [@Rothwell1995;@KRAVITZ2004]. Understanding
heterogeneity of treatment effects (HTE) has been crucial for both personalized
(or precision) medicine and comparative effectiveness research, giving rise to a
wide range of approaches for its discovery, evaluation and application in
clinical practice. A common approach to evaluating HTE in clinical trials is
through subgroup analyses, which are rarely adequately powered and can lead to
false conclusions of absence of HTE or exaggerate its presence
[@Hayward2006;@Kent2018]. In addition, patients differ with regard to multiple
characteristics simulatneously, resulting in much richer HTE compared to the one
explored with regular on-variable-at-a-time subgroup analyses [Kent, BMJ 2018].

Baseline risk is a summary score inherently related to treatment effect that
can represent more closely the variability in patient characteristics
[@Rothwell2005;@Hayward2006;@Kent2007;@Kent2008;@Kent2010]. For example, an
invasive coronary procedure---in comparison with medical treatment---improves
survival in patients with myocardial infarction at high (predicted) baseline
risk but not in those at low baseline risk [@Thune2005]. It has also been shown
that high-risk patients with pre-diabetes benefit substantially more from a
lifestyle modification program than low-risk patients [@Sussman2015].

Recently, systematic guidance on the application of risk-based methods for the
assessment of HTE has been developed for RCT data [@Kent2019;@PathEnE]. After
risk-stratifying patients using an existing or an internally derived prediction
model, risk stratum-specific estimates of relative and absolute treatment effect
are evaluated. Several methods for predictive HTE analysis have been adapted for
use in observational data, but risk-based methods are still not readily
available and have been highlighted as an important future research need
[@PathEnE].

The Observational Health Data Science and Informatics (OHDSI) collaborative has
established a global network of data partners and researchers that aim to bring
out the value of health data through large-scale analytics by mapping local
databases to the Observational Medical Outcomes Partnership (OMOP) Common Data
Model (CDM) [@hripcsak2015observational;@Overhage2012]. A standardized framework
applying current best practices for comparative effectiveness studies within the
OHDSI setting has been proposed [@Ryan2013]. This framework was successfully
implemented on a large scale for estimation of average effects of all first-line
hypertension treatment classes on a total of 52 outcomes of interest across a
global network of nine observational databases [@Suchard2019].

We aimed to develop a framework for risk-based assessment of treatment effect
heterogeneity in observational healthcare databases, extending the existing
methodology developed for the RCT setting. We implemented the framework in a
publicly available package providing an out-of-the-box solution for implementing
such analyses at scale within any observational database mapped to OMOP-CDM. In
a case study we analyzed heterogeneity of the effects of first-line hypertension
treatment.

@@

The proposed framework defines 5 distinct steps: 1) definition of the research
aim; 2) identification of the databases within which the analyses will be
performed; 3) prediction of outcomes of interest; 4) estimation of absolute and
relative treatment effects within risk strata; 5) presentation of the
results. We developed an open-source R-package for the implementation of the
proposed framework and made it publicly available
(https://github.com/OHDSI/RiskStratifiedEstimation). An overview of the entire
framework can be found in Figure \ref{fig:graphicalAbstract}.

@@

We use a comparative cohort design. This means that at least three cohorts of
patients need to be defined at this stage of the framework:

@@

Including in our analyses multiple databases representing the population of
interest potentially increases the generalizability of results. Furthermore, the
cohorts should preferably have adequate sample size with adequate follow-up time
to ensure precise effect estimation, even within smaller risk strata. Other
relevant issues such as the depth of data capture (the precision at which
measurements, lab tests, conditions are recorded) and the reliability of data
entry should also be considered.

@@

Our method relies on adequately separating patients into subgroups based on
their baseline risk for the outcomes of interest. Therefore, a model---either an
existing external model adequately validated on an internally developed
one---assigning patient-level risk is required. For internally developing a risk
prediction model we adopt the standardized framework focused on observational
data that ensures adherence to existing guidelines [@Reps2018; @Collins2015;
@Moons2015].

We first need to define a target cohort of patients, i.e. the set of patients on
whom the prediction model will be developed. In our case, the target cohort is
generated by pooling the already defined treatment and comparator cohorts. We
develop the prediction model on the propensity score-matched (1:1) subset of the
pooled sample to avoid differentially fitting between treatment arms, thus
introducing spurious interactions with treatment
[@Burke2014;@vanKlaveren2019]. We also need to define a set of patients that
experience the outcome of interest, i.e. the outcome cohort. Finally, we need to
decide the time frame within which the predictions will be carried out, i.e. the
patients' time at risk. Subsequently, we can develop the prediction model.

It is important that the prediction models display good discriminative ability
to ensure that risk-based subgroups are accurately defined. A performance
overview of the derived prediction models including discrimination and
calibration both in the propensity score matched subset, the entire sample and
separately for treated and comparator patients should also be reported.

@@

We estimate treatment effects (both on the relative and the absolute scale)
within risk strata defined using the prediction model of step 3. We often
consider four risk strata, but fewer or more strata can be considered depending on
the available power for accurately estimating stratum-specific treatment
effects. Effect estimation may be focused on the difference in outcomes for a
randomly selected person from the risk stratum (average treatment effect) or for
a randomly selected person from the treatment cohort within the risk stratum
receiving the treatment under study (average treatment effect on the treated).

@@

Prior to analyzing results, it is crucial to ensure that all diagnostics are passed
in all risk strata. The standard diagnostics we carry out include analysis of
the overlap of propensity score distributions and calculation of standardized
mean differences of the covariates before and after propensity score
adjustment. Finally, we use effect estimates for a large set of negative control
outcomes (i.e. outcomes known to not be related with any of the exposures under
study) to evaluate the presence of residual confounding not accounted for by
propensity score adjustment [@Schuemie2014;@Schuemie2571;@Schuemie2018].

@@

In the presence of a positive treatment effect and a well-discriminating
prediction model we expect an increasing pattern of the differences in the
absolute scale, even if treatment effects remain constant on the relative scale
across risk strata. Due to this scale-dependence of treatment effect
heterogeneity, results should be assessed both on the relative and the absolute
scale.

@@

As a demonstration, we evaluated if our proposed method was able to identify
treatment effect heterogeneity of ACE inhibitors compared to beta blockers using
acute myocardial infarction (MI) risk quarter specific effect estimates, both on
the relative and on the absolute scale. We focused on three efficacy outcomes
(acute MI, hospitalization with heart failure and ischemic or hemorrhagic
stroke) and six safety outcomes (hypokalemia, hyperkalemia, hypotension,
angioedema, cough and abnormal weight gain). We used data from three US-based
claims databases. The analysis plan was the framework outlined in steps 1
through 5.

@@

We considered the following research aim: “compare the effect  of ACE inhibitors
($T$) to the effect of beta blockers ($C$) in patients with established hypertension
with respect to nine outcomes ($O_1,\dots,O_9$)”. The cohorts are:

- Treatment cohort: Patients receiving any drug within the ACE inhibitor class
  with at least one year of follow-up before treatment initiation and a recorded
  hypertension diagnosis within that year.
- Comparator cohort: Patients receiving any drug within the beta blocker class
  with at least one year of follow-up before treatment initiation and a recorded
  hypertension diagnosis within that year.
- Outcome cohorts: We considered three efficacy and six safety outcome cohorts. These
  were patients in the database with a diagnosis of: acute MI; hospitalization
  with heart failure; ischemic or hemorrhagic stroke (efficacy outcomes);
  hypokalemia; hyperkalemia; hypotension; angioedema; cough; abnormal weight
  gain (safety outcomes). Among the safety outcomes we focus on angioedema and
  cough which are two known adverse events linked to treatment with ACE
  inhibitors [@Israili1992]. Results on the rest of the safety outcomes
  are included in the supplement.

@@

```{r, echo = FALSE}
incidenceOverall <- readRDS(here::here("data/processed/incidenceOverall.rds"))
patientNumbers <- incidenceOverall %>%
  group_by(database) %>%
  summarise(
    treatment = max(treatmentPersons),
    comparator = max(comparatorPersons)
  )
```

For our demonstration we used data from three US claims databases, namely IBM
MarketScan Commercial Claims and Encounters (CCAE), IBM MarketScan Medicaid
(MDCD), and IBM MarketScan Medicare Supplemental Beneficiaries (MDCR). Our
analyses included a total of
`r patientNumbers %>% filter(database == "ccae") %>% pull(treatment)`,
`r patientNumbers %>% filter(database == "mdcd") %>% pull(treatment)`, and
`r patientNumbers %>% filter(database == "mdcr") %>% pull(treatment)`
patients initiating treatment with ACE inhibitors and 
`r patientNumbers %>% filter(database == "ccae") %>% pull(comparator)`,
`r patientNumbers %>% filter(database == "mdcd") %>% pull(comparator)`, and
`r patientNumbers %>% filter(database == "mdcr") %>% pull(comparator)`
patients initiating treatment with beta blockers in CCAE, MDCD and MDCR
respectively (Table \@ref(tab:incidenceOverall)). Adequate numbers of patients were included in all strata
of predicted acute MI risk (Supplement: Table XX).

@@

We internally developed separate prediction models for acute MI in all three
databases. The prediction models were estimated on the propensity score matched
(1:1) subset of the sample, using caliper of 
`r format(.2, nsmall = 1, decimal.mark = intToUtf8("0x00B7"))`
and after excluding patients having the outcome any time prior to treatment
initiation. We chose a 2-year time at risk for patients and developed the
prediction models using LASSO logistic regression with 3-fold cross validation
for hyper-parameter selection.

The models had moderate discriminative performance (internally validated) with
no major issues of overfitting to any cohort except for the case of CCAE, where
the derived prediction model performed better in the comparator cohort (Table
\@ref(tab:predictionPerformance)). We also observed lower performance of the
prediction model developed in MDCR compared to the other 2 databases.

@@

We used patient-level predictions to stratify the sample into four acute MI risk
quarters. Within risk quarters, relative effects were estimated using Cox
regression and absolute effects were derived from the Kaplan-Meier estimate
differences at two years after treatment initiation. To adjust for observed
confounding within each risk quarter, we estimated propensity scores using the
same approach as step 3 and stratified patients into five propensity score
strata. The risk quarter-specific effect estimates were derived by averaging
over the estimates within the propensity score fifths.

In the lowest acute MI risk quarter of CCAE and MDCD we observed strong
separation of the propensity score distributions, therefore, effect estimates
derived in these strata are not well-supported (Figure
\ref{fig:psDensity}). This problematic behavior is also visible in the covariate
balance plots comparing standardized mean differences of patient characteristics
before and afrer PS adjustment, where in many cases the commonly accepted bound
of 
`r format(.1, nsmall = 1, decimal.mark = intToUtf8("0x00B7"))`
is violated (Figure \ref{fig:covariateBalance}). This is more pronounced in the
lowest acute MI risk quarter of CCAE, but remains an issue for a small number of
covariates in all CCAE risk strata. This diagnostic also fails for the two lower
acute MI risk quarters of MDCD. Often the persisting imbalances were linked to
pregnancy outcomes, which can be explained by the contraindication of ACE
inhibitors in this condition. Analyses in MDCR passed all diagnostics.

Finally, the distribution of the estimated relative risks with regard to 30
negative control outcomes indicated unresolved confounding within the lowest
acute MI risk quarter of CCAE (Figrue \ref{fig:negativeControls}). Hazard ratios
significantly different than 1 (true effect size) were concentrated in the lower
right part of Figure \ref{fig:negativeControls}: panel Q1. This suggests
significant negative effects of ACE inhibitors compared to beta blockers on
causally unrelated outcomes, pointing at unresolved differences between the two
treatment arms. This was not the case in the other risk quarters of CCAE, or in
any risk quarter of MDCD and MDCR (Supplement, Figure XX).

@@

```{r, echo=FALSE}
map_outcomes <- readRDS(here("data/processed/map_outcomes.rds"))

map_exposures <- readRDS(here("data/processed/map_exposures.rds"))
overall <- readRDS(here("data/processed/mappedOverallResults.rds")) %>%
  left_join(map_outcomes, by = c("outcomeId" = "outcome_id")) %>%
  select(-outcomeId) %>%
  rename("outcome" = "outcome_name") %>%
  left_join(map_exposures, by = c("treatmentId" = "exposure_id")) %>%
  select(-treatmentId) %>%
  rename("treatment" = "exposure_name") %>%
  left_join(map_exposures, by = c("comparatorId" = "exposure_id")) %>%
  select(-comparatorId) %>%
  rename("comparator" = "exposure_name") %>%
  mutate(
    outcome = gsub("_", " ", outcome),
    estimate = paste0(
      format(round(estimate, 2), nsmall = 2, decimal.mark = intToUtf8("0x00B7")),
      " (", 
      format(round(lower, 2), nsmall = 2, decimal.mark = intToUtf8("0x00B7")),
      ", ", 
      format(round(upper, 2), nsmall = 2, decimal.mark = intToUtf8("0x00B7")),
      ")"
    )
  ) %>%
  select(-c("treatment", "comparator", "analysisType", "seLogRr", "lower", "upper")) %>%
  relocate(c("database", "outcome"))

relative <- readRDS(here("data/processed/mappedOverallRelativeResults.rds")) %>%
  left_join(map_outcomes, by = c("estOutcome" = "outcome_id")) %>%
  select(-estOutcome) %>%
  rename("estOutcome" = "outcome_name") %>%
  left_join(map_outcomes, by = c("stratOutcome" = "outcome_id")) %>%
  select(-stratOutcome) %>%
  rename("stratOutcome" = "outcome_name") %>%
  left_join(map_exposures, by = c("treatment" = "exposure_id")) %>%
  select(-treatment) %>%
  rename("treatment" = "exposure_name") %>%
  left_join(map_exposures, by = c("comparator" = "exposure_id")) %>%
  select(-comparator) %>%
  rename("comparator" = "exposure_name")

absolute <- readRDS(here("data/processed/mappedOverallAbsoluteResults.rds")) %>%
  left_join(map_outcomes, by = c("estOutcome" = "outcome_id")) %>%
  select(-estOutcome) %>%
  rename("estOutcome" = "outcome_name") %>%
  left_join(map_outcomes, by = c("stratOutcome" = "outcome_id")) %>%
  select(-stratOutcome) %>%
  rename("stratOutcome" = "outcome_name") %>%
  left_join(map_exposures, by = c("treatment" = "exposure_id")) %>%
  select(-treatment) %>%
  rename("treatment" = "exposure_name") %>%
  left_join(map_exposures, by = c("comparator" = "exposure_id")) %>%
  select(-comparator) %>%
  rename("comparator" = "exposure_name") %>%
  mutate(
    estimate = 100 * estimate,
    lower    = 100 * lower,
    upper    = 100 * upper
  )

@@

acute_cough_absolute <- absolute %>%
  filter(
    stratOutcome == "acute_myocardial_infarction",
    estOutcome == "cough"
  )

acute_mi_stroke_relative <- relative %>%
  filter(
    stratOutcome == "acute_myocardial_infarction",
    estOutcome == "stroke"
  )

acute_mi_stroke_absolute <- absolute %>%
  filter(
    stratOutcome == "acute_myocardial_infarction",
    estOutcome == "stroke"
  )
```

The overall estimated hazard ratios for the main outcomes are presented in
Table \@ref(tab:overallTable). For hospitalization with acute MI there was an increasing trend in
favor ACE inhibitors compared to beta blockers on the relative scale (hazard
ratios decreased) with increasing acute MI risk. More specifically, hazard
ratios decreased from

@@

95\% CI) in CCAE and MDCD respectively (Figure 4). In MDCR hazard ratios
increased from

@@

95\% CI). Relative treatment effect estimates for hospitalization with heart
failure favored ACE inhibitors across all risk strata in all databases. In the
case of stroke in CCAE we found quite constant hazard ratios which became weaker
in the highest risk quarter patients
(`r acute_mi_stroke_relative %>% filter(database == "ccae", riskStratum == "Q4") %>% pull(estimate)`
with 95\% CI from 
`r acute_mi_stroke_relative %>% filter(database == "ccae", riskStratum == "Q4") %>% pull(lower)`
to
`r acute_mi_stroke_relative %>% filter(database == "ccae", riskStratum == "Q4") %>% pull(upper)`).
In the other two databases no significant relative treatment effects were
observed for stroke.  In terms of the safety outcomes, we found an increased ACE inhibitor
risk of cough and angioedema on the relative scale across all risk strata. In
the case of cough, this effect decreased with increasing risk of acute MI---from

@@

## Interpretation
